---
title: "Estimating Heritability and Genetic Covariance via LDSC"
author: "Yihe Yang (yxy1234@case.edu)"
date: ""
output: github_document
---

## Overview

This tutorial demonstrates how to perform single- and bivariate Linkage Disequilibrium Score Regression (LDSC) to estimate:

- SNP heritability of two behavioral traits: **driving time** and **computer use time**;
- Genetic and environmental covariance between them.

The example uses preprocessed GWAS summary statistics, originally derived from UK Biobank questionnaires, and LD scores based on European HapMap3 reference.

---

## Installation and Setup

```r
devtools::install_github("harryyiheyang/ldscR")
```

## Step 1: Load Example Data

```{r}
library(devtools)
library(dplyr)
library(ldscR)
data_url <- "http://tinyurl.com/nhdfwd8v"
temp_file <- tempfile()
download.file(data_url, temp_file, mode = "wb")
gwaslist <- readRDS(temp_file)
unlink(temp_file)

# Load reference alleles from HapMap3
data("hapmap3")
```

## Step 2: Align SNPs

```{r}
gwaslist <- filter_align(
  gwas_data_list = gwaslist,
  ref_panel = hapmap3[, c("SNP", "A1", "A2")]
)
```

## Step 3: Estimate SNP Heritability (Univariate LDSC)
```{r}
fit1 <- ldsc.univ(
  gwas = gwaslist$driving,
  LDSC = EURLDSC,
  nblock = 500,
  sampling.time = 200
)

fit2 <- ldsc.univ(
  gwas = gwaslist$computer,
  LDSC = EURLDSC,
  nblock = 500,
  sampling.time = 200
)

fit1  # Heritability and intercept for driving time
fit2  # Heritability and intercept for computer use time
```

## Step 4: Estimate Genetic Covariance (Bivariate LDSC)

```{r}
fit3 <- ldsc.bicov(
  gwas1 = gwaslist$driving,
  gwas2 = gwaslist$computer,
  LDSC = EURLDSC,
  nblock = 500,
  sampling.time = 200,
  h21 = fit1$h2,
  h22 = fit2$h2
)

fit3  # Genetic and sample-overlap covariance between traits
```

## Step 5: Construct Covariance Matrices

```{r}
gcov_matrix <- matrix(
  c(fit1$h2, fit3$gcov,
    fit3$gcov, fit2$h2),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("Driving", "Computer"), c("Driving", "Computer"))
)

ecov_matrix <- matrix(
  c(fit1$intercept, fit3$ecov,
    fit3$ecov, fit2$intercept),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("Driving", "Computer"), c("Driving", "Computer"))
)

gcov_matrix
ecov_matrix
```
